Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex

\lstset{basicstyle=\tt,numbers=none}
\begin{lstlisting}
  $\globalbarrier()$;
  if($\getgroupid()$ == 0) $\offerfork$();
  $\globalbarrier()$;
  $\offerkill()$;
  $\globalbarrier()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt,numbers=left}

The enclosing $\globalbarrier$ calls ensure that the change in number
of active workgroups from $M$ to $M'$ occurs entirely within the
resizing barrier, so that from a programmer's perspective, $M$ changes atomically.  The middle $\globalbarrier$ ensures that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.  That is, workgroups $[0, \textrm{min}(M, M') - 1]$ should be left intact.

Because $\resizingglobalbarrier$ can be implemented in the above
manner, we do not regard it \emph{conceptually} as a primitive of our
programming model.  However, we discuss in
\mysec\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
implemented significantly more efficiently if it is treated as a
primitive operation that interacts directly with the scheduler.

@@

We discuss the main semantic change induced by our programming
model---that the number of workgroups executing a kernel varies
dynamically---and illustrate the new language constructs and the
impact of this semantic change by adapting the work stealing and graph
traversal algorithms of \mysec\ref{sec:openclexamples} to use
cooperative kernels.

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it corresponds to the active group count $M$ and may change due to workgroups executing
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Being a function of $\getnumgroups()$, the value returned by $\getglobalsize()$
is also subject to change.
A cooperative kernel must therefore be written in a manner that is
robust to changes in the values returned by these functions.

In general, their volatility means that use of these functions should be avoided.
However, the situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
%
This guarantee can be exploited by algorithms that perform strided
data processing, which we illustrate below using the graph traversal
example.

>>>>

\lstset{basicstyle=\small\tt,numbers=none}
\begin{lstlisting}
  $\globalbarrier()$;
  if($\getgroupid()$ == 0) $\offerfork$();
  $\globalbarrier()$;
  $\offerkill()$;
  $\globalbarrier()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt,numbers=left}

The enclosing $\globalbarrier$ calls ensure that the change in number
of active workgroups from $M$ to $M'$ occurs entirely within the
resizing barrier, so that from a programmer's perspective, $M$ changes atomically.  The middle $\globalbarrier$ ensures that forking occurs before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are left intact.

Because $\resizingglobalbarrier$ can be implemented as above, we do not regard it \emph{conceptually} as a primitive of our
model.  However, in
\mysec\ref{sec:resizingbarrier} we show how a resizing barrier can be
implemented significantly more efficiently through direct interaction with the scheduler.

@@

%We discuss the main semantic change induced by our programming
%model---that the number of workgroups executing a kernel varies
%dynamically---and illustrate the new language constructs and the
%impact of this semantic change by adapting the work stealing and graph
%traversal algorithms of \mysec\ref{sec:openclexamples} to use
%cooperative kernels.

\paragraph{A changing workgroup count}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize()$
is similarly subject to change.
A cooperative kernel must thus be written in a manner that is
robust to changes in the values returned by these functions.

In general, their volatility means that use of these functions should be avoided.
However, the situation is more stable if a cooperative kernel does not call
$\offerkill$ and $\offerfork$ directly, so that only $\resizingglobalbarrier$ can affect the number of active
workgroups.  Then, at any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (considering the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The active workgroup count is constant within each resizing
barrier interval, so that 
$\getnumgroups()$ and $\getglobalsize()$ return stable values during such intervals.
%
As we illustrate below for graph traversal, this can be exploited by algorithms that perform strided
data processing.
